Global land characterisation using cover fractions at 100m resolution

### 2020-07-16 article/article.tex
Draft 2

% Minimalistic template for a generic article, geared towards MDPI guidelines
% Use PDFLaTeX
\documentclass[a4paper,10pt]{article}
% Packages provided by the MDPI template already
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.
\usepackage[sort&compress,sectionbib]{natbib}
\usepackage{graphicx}
\bibliographystyle{apalike}

% Custom packages
\usepackage[acronym,toc,shortcuts,nohypertypes=acronym]{glossaries}
\usepackage[colorlinks, allcolors=blue, unicode]{hyperref}
\usepackage{hyperxmp} % Import license metadata
\usepackage{titling} % Get the author for the metadata
\usepackage{textcomp} % \texttimes
%\usepackage[pdf]{graphviz} % Importing dot files; can be replaced with generated PDFs
\usepackage[norefs,nocites]{refcheck} % Develop: warn if figure is not referenced

% My own definition for the sections that start with a bold name but don't appear in TOC
\newcommand{\minisection}[1]{\medskip \textbf{#1:}}

% Metadata
\title{Global land cover fraction mapping: methods, variables, challenges}
\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Martin Herold, \\ Myroslava Lesiv, Marcel Buchhorn, Jan Verbesselt}
% * <Dainius Masiliūnas> 11:20:08 26 Mar 2019 UTC+0100:
% TODO: fix newline issue

\hypersetup{
    pdflicenseurl={http://creativecommons.org/licenses/by-sa/4.0/},
    pdfcopyright={This work is licensed under the Creative Commons Attribution-ShareAlike 4.0 International License.},
    pdfauthor={\theauthor}, % These are supposed to be the default but don't seem to be
    pdftitle={\thetitle},
    pdflang={en-GB}
}

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIRv}{\gls{NIR} of vegetation}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{Classification and regression tree}
\newacronym{GLM}{GLM}{General linear regression model}
\makenoidxglossaries
% Disable abbreviating RF for now [text={random forest}]
%\glsunset{RF}

@@

\maketitle

@@

\begin{itemize}
    \item 11 statistical and machine learning models tested for land cover fraction mapping
    \item Random forest regression performed the best, with RMSE of 17.3 globally
    \item A three-step regression method proposed, which results in the lowest MAE of 7.9
    \item Remote sensing data was the most important for model predictions, but all auxiliary data also contributed significantly
    \item Global wall-to-wall land cover fraction maps produced and their errors assessed
\end{itemize}

\abstract{Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a particular ratio. In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision, reduces legend complexity, and allows users the flexibility of defining their own mosaic discrete classes using thresholds they define themselves. To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.

We compared the performance of eight machine learning regression algorithms for land cover fraction mapping, for the best model assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.

Results showed that the three-model approach results in better estimation of extreme values (lower MAE of 8.4 vs 9.4) at the cost of misclassification (higher RMSE of 19.4 vs 17.3). From the compared models, random forest and cubist regressions reach the highest overall accuracy of 72\%, comparable to the reported accuracies of the current discrete global land cover maps. The models were driven primarily by spectral remote sensing data, followed by soil and climate data.

This research proves that machine learning algorithms can be applied globally to a wide variety of land cover to create an exhaustive global land cover fraction map. This expresses land cover more precisely, and empowers users to create their own discrete maps, which helps land cover monitoring and climate modelling communities. In addition, this work contributes to efforts to produce exhaustive global land cover fractions as an operational product, and to assess the accuracy of land cover fraction maps both thematically and spatially.

\minisection{Keywords} Fuzzy land cover classification, machine learning, random forest, gradient boosting, neural networks, fuzzy c-means, PROBA-V
}

\pagebreak
\tableofcontents
% * <Dainius Masiliūnas> 11:40:16 21 Apr 2020 UTC+0200:
% Temporary, to make it easier to understand the structure
\pagebreak

@@

% \begin{enumerate}
%  \item Relevance of fractional LC mapping (link to user requirements and SDGs)
%  \item Studies that have looked into fractional LC mapping so far
%  \item Available fractional methods that were introduced in literature (wavelet transform etc.)
%  \item Methods that are tested in this paper
%  \item Global land cover maps
% \end{enumerate}

Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \ac{CGLS-LC100} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

Most of these global land cover maps, as well as previous global land cover maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided as discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
This issue becomes worse at coarse resolutions and heterogeneous landscapes.
It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus ultimately lead to an underestimation of tree cover at large scales.

@@

Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor a global scale.

@@

All in all, with this study we aim to investigate and compare methods for performing land cover fraction mapping, at a global scale, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hamper land cover fraction mapping.
Therefore the objectives of our study are:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning models for a global-scale land cover fraction mapping task
 \item Investigate approaches for reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
 \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
\end{enumerate}

@@

For model validation, an independent global dataset of \ac{CGLS}, generated by Wageningen University \& Research \citep{tsendbazar_developing_2018}, was used.
The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.

The data includes over 150\,000 training points and over 20\,000 validation points across the globe, describing the fractions of 12 classes (see figure \ref{fig-reference-data}) as they were in the year 2015.
Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy a single pixel.
This means that for a given pixel, only a few of the 12 classes would normally have a fraction that is above 0\%.
This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
But for user communities, correct prediction of non-zero values is more important.

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/maps/2019-08-06-training-and-validation}
 \caption{Reference data points with the dominant land cover class. (a): training dataset. (b): validation dataset.}
 \label{fig-reference-data}
\end{figure}

@@

To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
In this study, six groups of covariates were used: spectral, harmonic, terrain, soil, climate and location (see table \ref{tab-inputdata}).
All of these covariates had to be preprocessed before they could be input into the models.

\begin{table}
\centering
    \begin{tabular}{llp{6cm}}
         \toprule
         \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
         \midrule
         Spectral & 22 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
         Harmonic & 9 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
         Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
         Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
         Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
         Location & 3 & None (intrinsic) \\
         \bottomrule
    \end{tabular}
    \label{tab-inputdata}
    \caption{Data sources for the covariates input into the model.}
\end{table}

\subsubsection{Optical covariates}
% * <Dainius Masiliūnas> 16:04:43 06 Aug 2019 UTC+0200:
% Maybe better to call it "Proba-V data" or such

The entire archive (2014-03-11 to 2019-07-16) of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{probavguide2} was used for this study.
The time series of NDVI and each of the bands at the points of interest were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were too far from the fitted \ac{LOESS} curve over the blue band.

The resulting cloud-free time series was then used to extract descriptive statistics (mean and interquartile range) of NDVI over the whole time series.
Next, harmonic analysis was run in order to decompose the time series into sine and cosine components for two frequencies (annual and semiannual), as well as the trend and intercept of the model.
From this, phase and amplitude for the two frequencies were calculated.

Lastly, extra \glspl{VI} (NDMI, EVI, OSAVI and NIR\textsubscript{v}) were calculated.
Each of these \glspl{VI} was then composited into a single image that represents the leaf-on season (three months) of 2015.

The optical preprocessing chain is displayed graphically in figure \ref{fig-preprocessing-optical}.
All of the outputs were then used in the processing chain.

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/flowcharts/preprocessing-optical}
 \caption{Optical data preprocessing chain.}
 \label{fig-preprocessing-optical}
\end{figure}

\subsubsection{Terrain covariates}

The ASTER GDEM v003 \citep{ASTGTM003} product was used as the elevation covariate.
In addition, terrain parameters were calculated out of elevation: slope, aspect and \ac{TPI}.
All terrain covariates were then resampled to the Proba-V grid.

\subsubsection{Climate covariates}

The climate data from WorldClim 2.0 30 s product \citep{worldclim2} was used.
Biophysical parameters were calculated from the data as well.
Some additional biophysical parameters were also calculated, namely the means, minimums and maximums over the whole time series of solar radiation, wind speed and water vapour pressure.

\subsubsection{Soil covariates}

Two products were used for data on soils: SoilGrids and LandGIS.
SoilGrids is based on a machine learning model that predicts soil properties at various soil depths globally at a 250 m resolution \citep{hengl_soilgrids250m_2017}
LandGIS is a successor to SoilGrids using a refined machine learning methodology \citep{hengl_predictive_2019}, however, it does not yet provide as many covariates as SoilGrids.

In the creation of SoilGrids, a land cover map (based on MODIS) was used.
In order to avoid circular dependencies, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
In addition, the taxonomy covariates were also excluded, since they are categorical data derived from the numerical soil property data.

\subsubsection{Other covariates}

The latitude, longitude and absolute latitude of the points were also included as location covariates.
Lastly, as part of the SoilGrids dataset, the MODIS land surface temperature product (MOD11A2, \citealp{wan_mod11a2_2015}) data monthly composites for 2011 were also used as additional covariates.

@@

Most of the colinear covariates were soil covariates at different depths.
Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
Similarly, climate data was colinear between subsequent months. Thus January and August data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.

While \ac{OSAVI} was calculated as a covariate, ultimately it was dropped due to too high correlation with \ac{NDVI}.
Similarly, the spectral bands of Proba-V were also highly correlated, so only the red and \ac{NIR} bands were kept.

@@

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/flowcharts/processing}
% * <Dainius Masiliūnas> 13:45:26 13 Aug 2019 UTC+0200:
% TODO: optimise style for readability.
 \caption{Processing chain.}
 \label{fig-processing}
\end{figure}

@@

For deriving land cover fractions, we compared a wide array of machine learning regression methods: \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression, \ac{MLR}, \ac{MLP} \glspl{NN}, \glspl{SVM}, Cubist and \ac{RF} regression.
In addition, as a baseline we also compare the results with an intercept model (all fractions always predicted to be equal).

@@

\subsubsection{\Glsentryfull{GLM}}

>>>>

% Use PDFLaTeX
\documentclass[review,authoryear,3p]{elsarticle}
\journal{Remote Sensing of Environment}
\bibliographystyle{model5-names}\biboptions{authoryear}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.
%\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
\usepackage{graphicx}
\usepackage{adjustbox} % For resizing tables
\usepackage{array} % For mid-alignment (m{xcm}) in tables
\usepackage{rotating} % Sideways figures
%\bibliographystyle{apalike} % elsarticle defined at the top

% Custom packages
\usepackage[acronym,toc,shortcuts,nohypertypes=acronym]{glossaries}
\usepackage[colorlinks, allcolors=blue, unicode]{hyperref}
%\usepackage{hyperxmp} % Import license metadata % Conflicts with elsarticle
%\usepackage{titling} % Get the author for the metadata % Conflicts with elsarticle
\usepackage{textcomp} % \texttimes
%\usepackage[pdf]{graphviz} % Importing dot files; can be replaced with generated PDFs
%\usepackage[norefs,nocites]{refcheck} % Develop: warn if figure is not referenced

% My own definition for the sections that start with a bold name but don't appear in TOC
\newcommand{\minisection}[1]{\paragraph{#1}}%{\medskip \textbf{#1:}}

% Metadata moved to frontmatter for elsarticle

% Requires hyperxmp and titling
% \hypersetup{
%     pdflicenseurl={http://creativecommons.org/licenses/by-sa/4.0/},
%     pdfcopyright={This work is licensed under the Creative Commons Attribution-ShareAlike 4.0 International License.},
%     pdfauthor={\theauthor}, % These are supposed to be the default but don't seem to be
%     pdftitle={\thetitle},
%     pdflang={en-GB}
% }

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{WUR}{WUR}{Wageningen University \& Research}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIRv}{\gls{NIR} of vegetation}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{Classification and regression tree}
\newacronym{GLM}{GLM}{General linear regression model}
\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
%\makenoidxglossaries
% Disable abbreviating RF for now [text={random forest}]
%\glsunset{RF}

@@

\begin{frontmatter} % elsarticle only

\title{Global land characterisation using cover fractions at 100m resolution}

\author[WURGRS]{Dainius Masiliūnas\corref{ca}}
\ead{dainius.masiliunas@wur.nl}
\cortext[ca]{Corresponding author}

\author[WURGRS]{Nandin-Erdene Tsendbazar}
\ead{nandin.tsendbazar@wur.nl}

\author[WURGRS]{Martin Herold}
\ead{martin.herold@wur.nl}

\author[IIASA]{Myroslava Lesiv}
\ead{lesiv@iiasa.ac.at}

\author[VITO]{Marcel Buchhorn}
\ead{marcel.buchhorn@vito.be}

\author[WURGRS]{Jan Verbesselt}
\ead{jan.verbesselt@wur.nl}

\address[WURGRS]{Wageningen University \& Research, Laboratory of Geo-Information Science and Remote Sensing, Droevendaalsesteeg 3, 6708 PB Wageningen, the Netherlands}
\address[IIASA]{International Institute for Applied Systems Analysis (IIASA), Schlossplatz 1, A-2361 Laxenburg, Austria}
\address[VITO]{Flemish Institute for Technological Research (VITO), Boeretang 200, BE-2400 Mol, Belgium}

\begin{abstract} % elsarticle: environment instead of a command \abstract{
Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a particular ratio. In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity. To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.

We compared the performance of three classes of machine learning regression algorithms for land cover fraction mapping, for the best model assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe at a sampling interval of 0.2 degrees.

Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%. Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved by a tuned random forest model with our proposed three-model approach, which results in better estimation of extreme values at the cost of more misclassification. The models were driven primarily by optical remote sensing data, followed by soil and climate data. Herbaceous cover was the most challenging to discern for all models, whereas built-up and water were the easiest to estimate.

This research proves that machine learning algorithms can be applied globally to a wide variety of land cover classes to create an exhaustive global land cover fraction map. This expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which helps to customise the result for a diverse range of uses. In addition, this work contributes to efforts to produce exhaustive global land cover fractions as an operational product, and to assess the accuracy of land cover fraction maps both thematically and spatially.
\end{abstract}

\begin{keyword}
Land cover fraction mapping\sep machine learning\sep random forest\sep Cubist\sep neural networks\sep fuzzy nearest centroid\sep lasso regression\sep PROBA-V
\end{keyword}

\end{frontmatter} % end elsarticle only

%\maketitle % elsarticle doesn't use it

@@

\begin{itemize}
    \item Three groups of statistical and machine learning models tested for deriving seven land cover fractions at a global scale
    \item Random forest regression performed the best (17.3 RMSE or 7.9 MAE)
    \item Remote sensing data was the most important for model predictions with varying features relevant for different fraction predictions
    \item Independent validation results show highest accuracies for built-up and water fractions, lower accuracies are found for grasses and shrubs
    \item Comprehensive multi-fraction global land cover characterisation shown as an alternative to discrete classification
\end{itemize}

%\pagebreak
%\tableofcontents
%\pagebreak

@@

Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received increasing amounts of attention, also as new satellites are launched and the capacity for land monitoring increases.
In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \ac{CGLS-LC100} map \citep{buchhorn_moderate_2019} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

All of these global land cover maps (with the exception of \ac{CGLS-LC100} fraction layers), as well as previous global land cover maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided as discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
This issue becomes worse at coarse resolutions and heterogeneous landscapes.
It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus ultimately lead to an underestimation of tree cover at large scales.

@@

Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor at global scale.

@@

Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
It requires a comprehensible dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
In this study, we made use of an independent validation set of over 20\,000 points collected across the globe using the same legend, grid and spatial resolution.
The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
We used the validation data to assess model accuracy both thematically and spatially.

All in all, with this study we aim to investigate and compare methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
Therefore the objectives of our study are:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping
 \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
 \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
 \item Assess the accuracy of the models from a thematic and spatial point of view, taking various user needs into account
\end{enumerate}

@@

For model validation, an independent global dataset of \ac{CGLS}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.

The data includes over 150\,000 training points and over 21\,000 validation points across the globe, describing the fractions of 12 classes in the year 2015.
The classification scheme follows the \gls{LCCS} \citep{buchhorn_moderate_2019}.
However, some of the classes were very detailed and had too few observations to classify well, thus we merged the classes into a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
The unknown class was discarded: points with the dominant land cover class marked as unknown were not used.
For points with a minority fraction of unknown, the remaining classes were linearly rescaled to add up to 100\%.
Thus in the end the training dataset size became 138\,164 and the validation dataset size became 20\,705. See figure \ref{fig-reference-data} for the spatial distribution of the points.

%Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy space within an area of a single pixel.
%This means that for a given pixel, only a few of the classes would normally have a fraction that is above 0\%.
%This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
%Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
%In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
%But for user communities, correct prediction of non-zero values is more important.

\begin{figure}
 \includegraphics[width=\textwidth]{article/article-figures/maps/2020-07-06-training-and-validation}
 \caption{Points representing 100x100 m areas at which land cover reference data was used in the study. The colours represent the dominant land cover class at each point. (a): training dataset, collected by \gls{IIASA}, 138\,164 points used in this study. (b): validation dataset, collected by \gls{WUR}, 20\,705 points used in this study.}
 \label{fig-reference-data}
\end{figure}

\begin{figure}
 \centering
 \includegraphics[width=0.75\textwidth]{article-figures/flowcharts/2020-07-10-flowchart}
 \caption{Processing workflow, from the raw input data to model accuracy assessment and wall-to-wall map output.}
 \label{fig-processing}
\end{figure}

@@

To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
See figure \ref{fig-processing} for an overview of the whole processing chain.

In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see table \ref{tab-inputdata}).
All of these covariates had to be preprocessed before they could be input into the models.

% \begin{table}
% \centering
%     \begin{tabular}{llp{6cm}}
%          \toprule
%          \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
%          \midrule
%          Spectral & 22 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
%          Harmonic & 9 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
%          Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
%          Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
%          Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
%          Location & 3 & None (intrinsic) \\
%          \bottomrule
%     \end{tabular}
%     \label{tab-inputdata}
%     \caption{Data sources for the covariates input into the model.}
% \end{table}

%\subsubsection{Vegetation indices and temporal metrics}
\label{sec-temporal-filter}

The entire archive (2014-03-11 to 2019-07-16) of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{probavguide2} was used for this study.
While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of Proba-V was chosen to obtain more robust statistics for the temporal metrics.
Land cover change is a relatively rare phenomenon and thus is not likely to affect more than 2\% of the training points within the time period.
Nevertheless, this may result in poorer model performance compared to if reference data for each year would be available.

The time series of each of the Proba-V bands were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the fitted \ac{LOESS} curve over the blue band.
The resulting cloud-free time series was then used to extract descriptive statistics (median and interquartile range) of NDVI, NDMI, EVI, OSAVI and NIR\textsubscript{v} over the whole time series, as well as for each season separately.
Next, harmonic analysis was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
From this, phase and amplitude for the two orders were calculated.
Lastly, the minimum and maximum of the NDVI time series was also retained as covariates.

% \begin{figure}
%  \includegraphics[width=\textwidth]{article-figures/flowcharts/preprocessing-optical}
%  \caption{Optical data preprocessing chain.}
%  \label{fig-preprocessing-optical}
% \end{figure}

%\subsubsection{Terrain covariates}

For elevation data, the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) was resampled to Proba-V 100 m grid and used as the elevation covariate.
In addition, GDAL was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
%All terrain covariates were then resampled to the Proba-V grid.

%\subsubsection{Climate covariates}

For climate data, WorldClim 2.0 30 s product \citep{worldclim2} was used.
It includes monthly temperature, precipitation, solar radiation, wind speed and water vapour pressure data.
In additional, 19 bioclimatic parameters were calculated from the data, using the dismo package \citep{dismo}.
Some additional biophysical parameters were also calculated, namely all of the climate data during the coldest, warmest, driest and wettest months of the year at each location, as well as yearly averages of each of the original parameters.

%\subsubsection{Soil covariates}

SoilGrids \citep{hengl_soilgrids250m_2017} was used for getting data about soils.
It is based on a random forest model that predicts soil properties at various soil depths globally at a 250 m resolution.
In the creation of SoilGrids, a land cover map (based on MODIS) had been used, and so, in order to avoid circular inference, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
The soil taxonomy covariates were also excluded, since they are categorical derivatives from the numerical soil property data and thus do not contribute to land cover fraction prediction.

%\subsubsection{Location covariates}

The latitude, longitude and absolute latitude of the reference points were also included when training the models, so that the models could learn spatial patterns.

@@

Most of the colinear covariates were soil covariates at different depths.
Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
Similarly, climate data was colinear between subsequent months. Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
In addition, all of the spectral bands of Proba-V were also highly correlated with each other as well as to vegetation indices, so only vegetation indices were kept.

@@

@@

For deriving land cover fractions, we compared a wide array of machine learning regression methods.
They can be broadly divided into three categories.
Tested linear models were \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression and \ac{MLR}.
Non-tree machine learning models tested were \ac{MLP} \glspl{NN} and \glspl{SVM}.
Lastly, machine learning models based on \glspl{CART} tested were Cubist and \ac{RF} regressions.
In addition, as a baseline we also compared the results with an intercept model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).

@@

\subsubsection{Linear models}
%\subsubsection{\Glsentryfull{GLM}}
